\speculake: Hiding Malware in Speculative Execution

### 2018-08-01 paper/related-work.tex
pass on A-C related work

\speculake shares many properties with \textit{weird machines}---a machine which
take advantage of bugs or unexpected idiosyncracies in existing systems to
perform arbitrary
computation~\cite{weird_machines,bratus2011exploit}. In
particular \speculake showcases the ability to execute dead code.

@@

Unexpected functionality such as instruction-less computing enables the
demonstration of a Turing complete weird machine by taking advantage of memory
translation tables, page-faults, and hardware exception handling to effectively
compute arithmetic and logical operations without the processor executing any
instructions~\cite{bangert2013page}.

@@

Previous ``weird machines'' have demonstrated the capability to execute
potentially malicious code in an unexpected manner, and \speculake is a new type
of ``weird machine'' that takes advantage of speculative execution and uses a
non-complicit trigger program to execute otherwise dead code.

\subsection{Red-Pills}
Hardware minutia have also been used to identify and fingerprint execution
environments---in malware this is know as a red-pill~\cite{red-pill} and is
typically used by malware to detect when it is under emulation or a debugging
environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
paleari2009fistful}

% Hardware bare-metal analysis

``Bare-metal'' emulation of malware~\cite{kirat2011barebox}, methods to
determine if malware is performing red-pill checks to avoid emulation
environments~\cite{kirat2014barecloud}, and symbolic execution can also be used
to combat red-pills by finding environmental triggers in malicious
binaries~\cite{schwartz2010all}.

Unlike other red-pills, \speculake confounds analysts and red-pill detectors
with the use a trigger program and speculative execution. The use of a separate
and benign trigger program controlled by the adversary enables the \speculake
malware to limit detection capabilities. Additionally, as \speculake is executed
via speculative execution, the instructions executed and speculative state are 
also obscured from analysts.

Similarly, debugging break-points will not enable analysts to debug \speculake
malware as break-points interrupt the control flow of a program and prevent
speculative execution. Symbolic execution will also fail to find the speculative
gadgets as it is never on the execution pathway. This allows \speculake to
effectively act as a red-pill while avoiding detection as such.

@@

There is a plethora of previous research into cache side-channels, examining how
information is leaked between
processes~\cite{percival2005cache,zhang2012cross,osvik2006cache}. This is a key
feature of Spectre~\cite{spectre} and Meltdown~\cite{meltdown}. In \speculake we
use the branch predictor side-channel as a \emph{covert
channel}~\cite{lampson1973note} between an innocent victim program and malware.
This allows \speculake to use the leaked information from cache as a message to
a colluding malware to make decisions.

@@

The branch predictor has been used as a covert
channel~\cite{evtyushkin2016understanding}. Previous work has examined the
capability of two colluding programs to share sensitive information between each
other using the branch predictor by timing how long a spy process executes taken
(not-taken) branches~\cite{evtyushkin2015covert}.

\speculake takes advantage of a cache side-channel in \texttt{OpenSSL} and uses
it as a covert channel. \texttt{OpenSSL} is not a malicious colluding program in
this case but rather a program leaking information on which cipher suite has
been selected by attackers, allowing us to use the branch predictor as an
attacker controlled covert channel of an innocent victim program.

>>>>

\speculake shares many properties with \textit{weird machines}---a machine which
takes advantage of bugs or unexpected idiosyncracies in existing systems to
perform arbitrary
computation~\cite{weird_machines,bratus2011exploit}. In
particular \speculake showcases the ability to execute dead code.

@@

Unexpected functionality such as instruction-less computing enables the
demonstration of a Turing complete ``weird machine'' by taking advantage of
memory translation tables, page-faults, and hardware exception handling to
effectively compute arithmetic and logical operations without the processor
executing any instructions~\cite{bangert2013page}.

@@

\speculake is a new type of ``weird machine'' that takes advantage of
speculative execution and uses a non-complicit trigger program to execute
otherwise dead code. While previous ``weird machines'' have demonstrated the
capability to execute potentially malicious code in an unexpected manner, we are
the first to exploit speculative execution in order to execute arbitrary
instructions.

\subsection{Red-Pills}
Hardware minutia have also been used to identify and fingerprint execution
environments---in malware this is known as a red-pill~\cite{red-pill} and is
typically used by malware to detect when it is under emulation or a debugging
environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
paleari2009fistful} attempting to circumvent malware detectors.

To address red-pills, systems often employ measures to detect them.
``Bare-metal'' emulation of malware~\cite{kirat2011barebox}, methods to
determine if malware is performing red-pill checks to avoid emulation
environments~\cite{kirat2014barecloud}, and symbolic execution can be used to
combat red-pills by finding environmental triggers in malicious
binaries~\cite{schwartz2010all}.

\speculake undertakes to avoid detection, and is a new class of red-pill. Unlike
previous red-pills, \speculake complicates analysis and impedes red-pill
detectors with the use a trigger program and speculative execution. The use of a
separate and seemingly benign trigger program influenced by the adversary
enables the \speculake malware to limit detection by acting benign until
triggered to influence the branch predictor. Additionally, as an \speculake
attack is carried out via speculative execution, the instructions executed and
speculative state are not committed to by the processor and are thus obscured
from analysts.

Similarly, break-points (and other debugging techniques) will not aid in
unravelling \speculake malware as break-points interrupt the control flow of a
program and prevent speculative execution. Symbolic execution will also fail to
find the speculative gadgets as they are never on the (committed) execution
pathway. This allows \speculake to effectively act as a red-pill while avoiding
detection as such.

@@

There is a plethora of previous research into cache side-channels, examining how
information is leaked between
processes~\cite{percival2005cache,zhang2012cross,osvik2006cache}. This is a key
feature of Spectre~\cite{spectre} and Meltdown~\cite{meltdown}. In \speculake we
use the branch predictor side-channel as a \emph{covert
channel}~\cite{lampson1973note} between an innocent trigger program and malware.
This allows \speculake to use the leaked information from cache as a message to
the malware in order to influence speculative execution pathways.

@@

The branch predictor has also been used as a covert
channel~\cite{evtyushkin2016understanding}. Previous work has examined the
capability of two colluding programs to share sensitive information between each
other using the branch predictor by timing how long a spy process executes taken
(not-taken) branches~\cite{evtyushkin2015covert}.

\speculake exemplifies a new covert channel using the branch predictor in which
only one side of the channel is nefarious and the other is innocent. \speculake
takes advantage of a branch predictor side-channel in \texttt{OpenSSL} and uses
it as a covert channel. \texttt{OpenSSL} is not a malicious colluding program in
this case but rather a program whose execution pathway can be influenced by
attackers based on selected cipher suites. This allows malware to use the branch
predictor as an attacker controlled covert channel between an unsuspecting
trigger program and \speculake malware.
